Fast approximate Bayesian inference of HIV indicators using PCA adaptive Gauss-Hermite quadrature

### 2023-07-18 src/docs_paper/paper.Rmd
Eventually make map figure with PCA-AGHQ rather than TMB #55

Let $\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA}) = - \partial^2 \log p_\texttt{LA}(\hat{\btheta}_\texttt{LA}, \y)$ be the curvature at the mode $\hat{\btheta}_\texttt{LA}$ and $[\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})]^{-1} = \hat{\mathbf{P}}_\texttt{LA} {\hat{\mathbf{P}}_\texttt{LA}}^\top$ be a matrix decomposition of the inverse curvature, then an AGHQ estimate of the normalising constant $p(\y)$ based on the Laplace approximation is
\begin{equation}
p(\y) \approx \int_{\btheta} \tilde p_\texttt{LA}(\btheta, \y) \approx \tilde p_\texttt{AGHQ}(\y) = |\hat{\mathbf{P}}_\texttt{LA}|\sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y) \omega(\z). \label{eq:aghq}
\end{equation}
The unadapted nodes are shifted by the mode and rotated by a decomposition of the inverse curvature such that $\z \mapsto \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}$.
Two alternatives [@jackel2005note] are

1. the Cholesky decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{L}}_\texttt{LA}$, where $\hat{\mathbf{L}}_\texttt{LA}$ is lower triangular, and
2. the spectral decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{E}}_\texttt{LA} \hat{\mathbf{\Lambda}}_\texttt{LA}^{1/2}$, where $\hat{\mathbf{E}}_\texttt{LA} = (\hat{\mathbf{e}}_{\texttt{LA}, 1}, \ldots \hat{\mathbf{e}}_{\texttt{LA}, m})$ contains the eigenvectors of $[\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})]^{-1}$ and $\hat{\mathbf{\Lambda}}_\texttt{LA}$ is a diagonal matrix containing its eigenvalues $(\hat \lambda_{\texttt{LA}, 1}, \ldots, \hat \lambda_{\texttt{LA}, m})$.

Equation \ref{eq:aghq} may then be used to normalise the Laplace approximation
\begin{equation}
\tilde p_\texttt{LA}(\btheta \, | \, \y) = \frac{\tilde p_\texttt{LA}(\btheta, \y)}{\tilde p_\texttt{AGHQ}(\y)}.
\end{equation}
To obtain inferences for the latent field (Equation \ref{eq:inla1}) we reuse the adapted nodes and weights [@rue2009approximate; @stringer2022fast]
\begin{equation}
\tilde p(\x \, | \, \y) = |\hat{\mathbf{P}}_\texttt{LA}| \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\texttt{G}(\x \, | \, \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y) \tilde p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA} \, | \, \y) \omega(\z). \label{eq:nest}
\end{equation}
Samples from this mixture of Gaussians may be obtained by drawing a node $\z$ with multinomial probabilities $\lambda(\z) = |\hat{\mathbf{P}}_\texttt{LA}| p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA} \, | \, \y) \omega(\z)$, then drawing from the corresponding Gaussian $\tilde p_\texttt{G}(\x \, | \, \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y)$.

@@

Our goal was to determine the accuracy of the approximate methods (TMB and PCA-AGHQ) as compared with the gold-standard (NUTS).
Settings used for each inferential method are provided in Table \ref{tab:inference-methods}, and, where relevant, discussed further below.
The `TMB` C++ user-template used to specify the log-posterior was the same for each approach.
The dimension of the latent field was $N = 467$ and the dimension of the hyperparameters was $m = 24$.
For the deterministic methods, following inference we simulated hyperparameter and latent field samples.
For all methods, we simulated age-sex-district specific HIV prevalence, ART coverage and HIV incidence from the latent field and hyperparameter posteriors.
<!-- Note: eventually this figure should be generated using the algorithm proposed in the paper -->
To provide intuition, model outputs from TMB are illustrated in Figure \ref{fig:naomi-results}.

```{r naomi-results, fig.cap="District-level model outputs for adults aged 15-49. Inference conducted using TMB."}
knitr::include_graphics("depends/naomi-results.png")
```

@@

df_point_pct <- df_point %>%
  ungroup() %>%
  group_by(indicator) %>%
  summarise(
    rmse_diff = 100 * diff(rmse) / max(rmse),
    mae_diff = 100 * diff(mae) / max(mae)
  )

rmse_ahgq_mean <- filter(df_point, method == "PCA-AGHQ", indicator == "Posterior mean estimate") %>% pull(rmse) %>% signif(2)
rmse_tmb_mean <- filter(df_point, method == "TMB", indicator == "Posterior mean estimate") %>% pull(rmse) %>% signif(2)
rmse_diff_mean <- filter(df_point_pct, indicator == "Posterior mean estimate") %>% pull(rmse_diff) %>% signif(2)

rmse_ahgq_sd <- filter(df_point, method == "PCA-AGHQ", indicator == "Posterior SD estimate") %>% pull(rmse) %>% signif(2)
rmse_tmb_sd <- filter(df_point, method == "TMB", indicator == "Posterior SD estimate") %>% pull(rmse) %>% signif(2)
rmse_diff_sd <- filter(df_point_pct, indicator == "Posterior SD estimate") %>% pull(rmse_diff) %>% signif(0)
```

@@

```{r ks, fig.cap="Example KS test for one parameter."}
knitr::include_graphics("figB.png")
```

>>>>

Let $\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA}) = - \partial^2 \log p_\texttt{LA}(\hat{\btheta}_\texttt{LA}, \y)$ be the curvature at the mode $\hat{\btheta}_\texttt{LA}$ and $[\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})]^{-1} = \hat{\mathbf{P}}_\texttt{LA} {\hat{\mathbf{P}}_\texttt{LA}}^\top$ be a matrix decomposition of the inverse curvature, then an AGHQ estimate of the normalising constant $p(\y)$ based on the Laplace approximation is
\begin{equation}
p(\y) \approx \int_{\btheta} \tilde p_\texttt{LA}(\btheta, \y) \approx \tilde p_\texttt{AGHQ}(\y) = |\hat{\mathbf{P}}_\texttt{LA}|\sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y) \omega(\z). \label{eq:aghq}
\end{equation}
The unadapted nodes are shifted by the mode and rotated by a decomposition of the inverse curvature such that $\z \mapsto \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}$.
Two alternatives [@jackel2005note] are
1. the Cholesky decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{L}}_\texttt{LA}$, where $\hat{\mathbf{L}}_\texttt{LA}$ is lower triangular, and
2. the spectral decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{E}}_\texttt{LA} \hat{\mathbf{\Lambda}}_\texttt{LA}^{1/2}$, where $\hat{\mathbf{E}}_\texttt{LA} = (\hat{\mathbf{e}}_{\texttt{LA}, 1}, \ldots \hat{\mathbf{e}}_{\texttt{LA}, m})$ contains the eigenvectors of $[\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})]^{-1}$ and $\hat{\mathbf{\Lambda}}_\texttt{LA}$ is a diagonal matrix containing its eigenvalues $(\hat \lambda_{\texttt{LA}, 1}, \ldots, \hat \lambda_{\texttt{LA}, m})$.
Equation \ref{eq:aghq} may then be used to normalise the Laplace approximation
\begin{equation}
\tilde p_\texttt{LA}(\btheta \, | \, \y) = \frac{\tilde p_\texttt{LA}(\btheta, \y)}{\tilde p_\texttt{AGHQ}(\y)}.
\end{equation}
To obtain inferences for the latent field (Equation \ref{eq:inla1}) we reuse the adapted nodes and weights [@rue2009approximate; @stringer2022fast]
\begin{equation}
\tilde p(\x \, | \, \y) = |\hat{\mathbf{P}}_\texttt{LA}| \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\texttt{G}(\x \, | \, \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y) \tilde p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA} \, | \, \y) \omega(\z). \label{eq:nest}
\end{equation}
Samples from this mixture of Gaussians may be obtained by drawing a node $\z$ with multinomial probabilities $\lambda(\z) = |\hat{\mathbf{P}}_\texttt{LA}| p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA} \, | \, \y) \omega(\z)$, then drawing from the corresponding Gaussian $\tilde p_\texttt{G}(\x \, | \, \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y)$.

@@

Our goal was to determine the accuracy of the approximate methods (TMB and PCA-AGHQ) as compared with the gold-standard (NUTS).
Settings used for each inferential method are provided in Table \ref{tab:inference-methods}, and, where relevant, discussed further below.
The `TMB` C++ user-template used to specify the log-posterior was the same for each approach.
The dimension of the latent field was $N = 467$ and the dimension of the hyperparameters was $m = 24$.
For the deterministic methods, following inference we simulated hyperparameter and latent field samples.
For all methods, we simulated age-sex-district specific HIV prevalence, ART coverage and HIV incidence from the latent field and hyperparameter posteriors.
To provide intuition, model outputs from TMB are illustrated in Figure \ref{fig:naomi-results}.

```{r naomi-results, fig.cap="District-level model outputs for adults aged 15-49. Inference conducted using TMB."}
knitr::include_graphics("depends/naomi-results.png")

# Eventually this figure should be generated using PCA-AGHQ
# knitr::include_graphics("depends/figB.png")
```

@@

df_point_pct <- df_point %>%
  ungroup() %>%
  group_by(indicator, type) %>%
  summarise(
    rmse_diff = 100 * diff(rmse) / max(rmse),
    mae_diff = 100 * diff(mae) / max(mae)
  )

rmse_ahgq_mean <- filter(df_point, method == "PCA-AGHQ", indicator == "Posterior mean estimate", type == "latent") %>% pull(rmse) %>% signif(2)
rmse_tmb_mean <- filter(df_point, method == "TMB", indicator == "Posterior mean estimate", type == "latent") %>% pull(rmse) %>% signif(2)
rmse_diff_mean <- filter(df_point_pct, indicator == "Posterior mean estimate", type == "latent") %>% pull(rmse_diff) %>% signif(2)

rmse_ahgq_sd <- filter(df_point, method == "PCA-AGHQ", indicator == "Posterior SD estimate", type == "latent") %>% pull(rmse) %>% signif(2)
rmse_tmb_sd <- filter(df_point, method == "TMB", indicator == "Posterior SD estimate", type == "latent") %>% pull(rmse) %>% signif(2)
rmse_diff_sd <- filter(df_point_pct, indicator == "Posterior SD estimate", type == "latent") %>% pull(rmse_diff) %>% signif(0)
```

@@

```{r ks, fig.cap="Example KS test for one parameter."}
knitr::include_graphics("figC.png")
```
